Space for WHAM: a multi-region, multi-stock generalization of the Woods Hole Assessment Model with an application to black sea bass

### 2025-01-02 paper/paper.Rmd
saving work

A state-space statistical approach and maximum marginal likelihood or Bayesian fitting of stock assessment models allows estimation of time- and age-varying population attributes as random effects [@nielsenberg14,@cadigan16,@milleretal16]. This estimation approach is considered an essential feature of gold-standard assessment models that we use in tactical management of commercially important fish stocks [@puntetal20]. The State-space Assessment Model [SAM, @nielsenberg14] continues to be developed and remains widely used within ICES to assess European fish stocks. Various state-space models are being used to manage cod and plaice stocks in the waters of Eastern Canada [@varkeyetal22;@perreaultetal20] and to the south, the Woods Hole Assessment Model [WHAM, @stockmiller21] is now used to assess many fish stocks in the Northwest Atlantic Ocean [@nefsc22;@nefsc22a;@nefsc24].

WHAM is an R package developed and maintained at NOAA's Northeast Fisheries Science Center [https://timjmiller.github.io/wham, @millerstock20; @stockmiller21]. WHAM can be configured to fit a wide range of age-structured models from traditional statistical catch-at-age models without any random effects to models with several time and age varying process errors and possibly effects of environmental covariates on various demographic parameters. Like SAM, WHAM models are built using the Template Model Builder package [TMB, @kristensenetal16] which provides a computationally efficient means of fitting an extremely wide class of models with random effects. WHAM has undergone active development since its creation and includes random effects options for in recruitment, inter-annual transitions numbers at age (hereafter referred to as "survival"), fishery and index selectivity, natural mortality, and catchability.

However, WHAM has up to now only allowed models with one stock (component) and a single region and without seasonal changes in stock dynamics. Using such models for stocks that have subcomponents with varying seasonal movement can provide incorrect inferences and poor management advice [@bosleyetal22;@yingetal11;@caoetal14]. Furthermore, the ability to account for spatial structure and model multiple stocks are also important features of leading-edge assessment modeling frameworks [@puntetal20]. We describe here the implementation of these features and other extensions since @stockmiller21 in WHAM version 2.0. Many of these new configuration options can be useful whether modeling 1 or more stocks and regions.

This extension of WHAM was developed in concert with a stock assessment for black sea bass through the NEFSC research track assessment process where new modeling frameworks were recently examined and modeling multiple stocks or stock components simultaneously was of interest [@nefsc23]. \textcolor{red}{[Text on black sea bass here]} We apply WHAM 2.0 to two stock components of black sea bass off the coast of the NEUS and evaluate evidence for alternative hypotheses of temporal variation and effects specifically of bottom temperature on recruitment and natural mortality of age 1 individuals.

@@

Many of the options and equations of WHAM version 2.0 are the same as those in @stockmiller21, so we will only describe extensions and differences that have occurred since their first description of WHAM. The new version of WHAM can model multiple stocks and survival, movement, harvest and natural mortality are tracked for each stock. Much of the description below is for a specific stock $s$, but, for simplicity, this subscript is implicit except when necessary.

@@

As in previous versions of WHAM, the transitions in numbers at age from one year to another after recruitment can be treated deterministically or as functions of random effects. The predicted numbers at age in year $y$ at age $a$ for a given stock are vector analogs ($\mathbf{N}_{O,y,a}$) of the equations for numbers at age in the standard WHAM model [@stockmiller21]. For ages $a = 2,\ldots, A-1$, where $A$ is the plus group, the expected number alive in each region at the beginning of the following year and next age class age can be obtained from the first $n_R$ elements of the vector
$$\mathbf{P}_{y-1,a-1}' \mathbf{N}_{y-1,a-1}.$$
The numbers alive in each region can also be modeled more simply using the sub-matrix $\mathbf{O}_{y,a}$. The general model for the transitions in abundance at age is 
\begin{equation*}
\log\left(\mathbf{N}_{O,y,a}\right)|\mathbf{N}_{O,y-1,a-1} =  \log\left(\mathbf{O}_{y-1,a-1}' \mathbf{N}_{O,y-1,a-1}\right) + \boldsymbol{\varepsilon}_{y,a}
\end{equation*}
for ages $a = 2,\ldots, A-1$, and for the plus group
\begin{equation*}
\log\left(\mathbf{N}_{O,y,A}\right)|\mathbf{N}_{O,y-1,a-1},\mathbf{N}_{O,y-1,A} = \log\left(\mathbf{O}_{y-1,A-1}' \mathbf{N}_{O,y-1,A-1} + \mathbf{O}_{y-1,A}' \mathbf{N}_{O,y-1,A}\right) + \boldsymbol{\varepsilon}_{y,A}.
\end{equation*}
When the transitions in abundance at age are treated deterministically, $\boldsymbol{\varepsilon}_{y,a} = 0$. The stock- and region-specific errors $\boldsymbol{\varepsilon}_{y,a}$ are independent, but the same correlation structures as previous versions are possible across ages and years for a given stock and region. When there is autocorrelation with age, WHAM now assumes this applies only to ages $a>1$ by default so that recruitment random effects are independent of those for for the annual transitions of older age classes. So the general covariance structure for a given stock at ages $a>1$ in region $r$ is that of a two-dimensional first-order autoregressive (2DAR1) process
\begin{equation*}
  Cov\left(\epsilon_{y,a,r},\epsilon_{y',a',r}\right) =   \frac{\rho_{N,\text{age},r}^{|a-a'|}\rho_{N,\text{year},r}^{|y-y'|}\sigma_{N,a,r}\sigma_{N,a',r}}{\left(1 -  \rho_{N,\text{age},r}^2\right)\left(1 - \rho_{N,\text{year},r}^2\right)} 
\end{equation*}
and that for age 1 is just AR1 across years 
\begin{equation*}
  Cov\left(\epsilon_{y,1},\epsilon_{y',1}\right) =   \frac{\rho_{N,1,\text{year}}^{|y-y'|}\sigma^2_{N,1}}{1 - \rho_{N,1,\text{year}}^2}.
\end{equation*}
Since recruitment for a given stock currently only occurs in one region $r_s$ there is only a single time-varying recruitment random effect for each stock.

@@

For each season, there are at most $n_R-1$ parameters determining movement among regions given starting an the season in region $r$ in either the sequential or simultaneous configurations. Movement parameters are estimated on a transformed scale via a link function $g(\cdot)$. If survival and movement are occur simultaneously, the parameters are estimated a log link function is used and if they are separable, an additive logit link function (like a multinomial regression) is used. On the transformed scale, the general model for the movement parameter from region $r$ to $r'$ in season $i$ and year $y$ for individuals of age $a$ is a linear function of both random and environmental effects:
\begin{equation*}
  g(\mu_{r\rightarrow r',y,a,i}) = \theta_{r\rightarrow r',i} + \epsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r \rightarrow r',a,i,k} E_{k,y}.
\end{equation*}
The random effects $\epsilon_{r\rightarrow r',y,a,i}$ are season-, and(or) region-to-region-specific and modeled most generally as 2DAR1 random effects with age and(or) year where the covariance is
\begin{equation*}
  Cov\left(\epsilon_{r\rightarrow r',y,a,i},\epsilon_{r\rightarrow r',y',a',i}\right) =   \frac{\rho_{r\rightarrow r',\text{age},i}^{|a-a'|}\rho_{r\rightarrow r',\text{year},i}^{|y-y'|}\sigma^2_{r\rightarrow r',i}}{\left(1 -  \rho_{r\rightarrow r',\text{age},i}^2\right)\left(1 - \rho_{r\rightarrow r',\text{year},i}^2\right)}
\end{equation*}
similar to how WHAM models variation in survival, natural mortality, and selectivity. Effects of covariate $E_k$ can be age-, season-, and(or) region-to-region-specific $\beta_{r\rightarrow r',a,i,k}$ and the same orthogonal polynomial options in the previous versions of WHAM for effects on recruitment and natural mortality are available.

There is currently no likelihood component for tagging data. Therefore, movement parameters would generally either need to be fixed or assumed to have some prior distribution, possibly based on external parameter estimates. We include prior distributions for the season and region-to-region specific (mean) movement parameters which are treated as random effects with the mean defined by the initial value of the fixed effect counterpart and standard deviation
  \begin{equation*}
  \gamma_{r\rightarrow r',i} \sim \text{N}\left(\theta_{r\rightarrow r',i}, \sigma^2_{r\rightarrow r',i}\right).
  \end{equation*}
When priors are used, the movement is defined instead as
  \begin{equation*}
  g(\mu_{r\rightarrow r',y,a,i}) = \gamma_{r\rightarrow r',i} + \epsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r\rightarrow r',a,i,k} E_{k,y} 
  \end{equation*}

@@

Natural mortality options have been expanded in WHAM. When not estimated, (mean) mortality rates may be stock-, region-, and age-specific. When random effects are used, the same 2DAR1 structure with age and year as described by @stockmiller21 can be configured for a given stock and region. Any environmental covariate effects can be stock-, region-, and age-specific. So the general model for natural mortality is
\begin{equation*}
  \log M_{y,a,r} = \theta_{M,r} + \epsilon_{M,r,y,a} + \sum^{n_E}_{k=1} \beta_{M,r,a,k} E_{k,y}.
\end{equation*}
The general covariance structure for random effects are modeled most generally as 2DAR1 random effect with age and(or) year where the covariance is
\begin{equation*}
  Cov\left(\epsilon_{M,y,a,r},\epsilon_{M,y',a',r}\right) =   \frac{\rho_{M,\text{age},r}^{|a-a'|}\rho_{M,\text{year},r}^{|y-y'|}\sigma^2_{M,r}}{\left(1 -  \rho_{M,\text{age},r}^2\right)\left(1 - \rho_{M,\text{year},r}^2\right)}.
\end{equation*}

@@

Catchability of the index can also be treated as functions of normal random effects and(or) environmental covariate effects 
$$log \frac{q_{m,y}-l_m}{u_m-q_{m,y}} = \theta_{q,m} + \varepsilon_{q,m,y}  + \sum^{n_E}_{k=1} \beta_{q,m,k} E_{k,y}$$ 
where $u_{m}$ and $l_{m}$ are the upper and lower bounds of catchability for index $m$ (defaults are 0 and 1000) and the general covariance structure for the annual random effects is AR1
$$Cov\left(\epsilon_{q,m,y},\epsilon_{q,m,y'}\right) =   \frac{\rho_{q,m}^{|y-y'|}\sigma^2_{q,m}}{1 - \rho_{q,m}^2}.$$

@@

For $X\%$ SPR-based reference points, we use a Newton method and iterate
\begin{equation}\label{eq:newton}
  \log\widetilde{F}^{(i)} = \log\widetilde{F}^{(i-1)} - \frac{g\left(\log\widetilde{F}^{(i-1)}\right)}{g'\left(\log\widetilde{F}^{(i-1)}\right)}
\end{equation}
where $g(\log F)$ is the difference between the weighted sums of spawning biomass per recruit at $F$ and $X$\% of unfished spawning biomass per recruit across stocks:
\begin{equation*}
  g(\log F) = \sum^{n_s}_{s=1} \lambda_s\left[\upphi_s\left(F = e^{\log F}\right) - \frac{X}{100}\upphi_s\left(F=0\right)\right].
\end{equation*}
where $\upphi_s\left(F=0\right)$ is the equilibrium unfished spawning biomass per recruit. $g'(\log F)$ is the derivative of $g$ with respect to $\log F$, and the weights to use for each stock $\lambda_s$ can be specified by the user or relative to the average of recruitment for each stock over the same years the user defines to calculate ``static'' equilibrium spawning biomass and yield.

@@

The projection options are generally the same as those for previous versions of WHAM. When there is movement of any stocks, the user has the option to project and use any random effects for time-varying movement or use the average over user specified years, analogous to how natural mortality can be treated in the projection period. The projection of any environmental covariates has been revised to better include error in the estimated latent covariate in any effects on the population in projection years.

@@

@@

Leading up to the 2023 peer-reviewed assessment, a working group (hereafter referred to as "working group") composed of scientists from federal, state, and academic institutions determined an optimal data and model configuration for the black sea bass stock using the multi-stock and multi-region extension of WHAM described above [@nefsc23]. This assessment included the spatial features and investigated inclusion of hypothesized environmental drivers that were prioritized research recommendations from previous black sea bass assessments.

Below we describe the assumptions and configuration of the assessment model as determined by the working group as well as the alternative assumptions for recruitment and natural mortality in the models we fit to evaluate alternative hypotheses of bottom temperature effects on black sea bass.

@@

The two monthly movement matrices are
\begin{equation*}
\boldsymbol{\mu}_{1} = 
  \begin{bmatrix}
     1-\mu_{\text{N}\rightarrow \text{S}} & \mu_{\text{N}\rightarrow \text{S}} \\
     0 & 1 \\
  \end{bmatrix}
\end{equation*}
for the portion of the year after spawning and
\begin{equation*}
\boldsymbol{\mu}_{2} = 
  \begin{bmatrix}
     1 &  0 \\
     \mu_{\text{S}\rightarrow \text{N}} & 1-\mu_{\text{S}\rightarrow \text{N}} \\
  \end{bmatrix}
\end{equation*}
for the portion of the year before spawning. As noted in the description of the general WHAM model, tagging data are not yet allowed. However, the working group also fit a Stock Synthesis model [@methotwetzel13] which provided estimated movement rate parameters and standard errors that were used to configure the priors for WHAM (see Supplementary Materials).

@@

## Uncertainty recreational CPA index observations {-}

The estimated coefficients of variation (CVs) provided by the analyses to generate the recreational catch per angler (recreational CPA) index ranged between 0.02 and 0.06 which the working group felt did not capture the true observation uncertainty in the index with regard to its relationship to stock abundance. In many of the initial runs as well as the base model we allowed a scalar multiple of the standard deviation of the log aggregate index to be estimated for these indices in the northern and southern regions. Models that successfully estimated these scalars indicated standard deviations for these surveys to be  approximately 5 times the input value and this value was fixed in many preliminary runs to avoid dealing with convergence problems. However, the base model successfully estimated these scalars. The model estimates are negligibly affected by estimating these scalars, but we felt estimating these parameters allowed uncertainty in model output to be more properly conveyed.

@@

The working group investigated many alternative assumptions for the probability models and selectivity models for the 8 different sets of age composition observations to reduce residual patterns and retrospective patterns. These analyses resulted in use of selectivity random effects for the northern fleet and indices and logistic-normal likelihoods for 6 sets of of age composition observations and Dirichlet-multinomial likelihoods for one index and one fleet in the northern region (Table \ref{tab:age-comp-sel-table}).

@@

\begin{equation}\label{eq:expected-recruitment}
E\left(\log N_{y,1}|x_y\right) = \mu_{R} + \beta_{R} x_y.
\end{equation}
Similarly, expected log-natural morality as a function of bottom temperature is
\begin{equation}\label{eq:expected-M1}
E\left(\log M_{y,1}|x_y\right) = \mu_{M,1} + \beta_{M} x_y
\end{equation}
Because age 1 fish for the northern component can exist in both regions after January 1, natural mortality is acting in both regions for this stock. For models with covariate and/or annual random effects for age 1 fish we assume them only in the northern regions for the northern component. The corresponding random effects are
\begin{equation}\label{eq:Rec-re}
\log N_{y,1} = E\left(\log N_{y,1}|x_y\right) + \epsilon_{y,1}
\end{equation}
and
\begin{equation}\label{eq:M-re}
\log M_{y,1} = E\left(\log M_{y,1}|x_y\right) + \epsilon_{M,y}.
\end{equation}
The constant or mean log-natural mortality rate is assumed to be $\mu_{M,a} = \log(0.4)$ for all ages as recommended by the working group. Because the bottom temperature anomalies and random effects are centered at 0, mean log-natural mortality at age 1 over the time series should be approximately equal to $\mu_{M,1}$ for models that include those effects.

## Model fitting, diagnostics, and projections {-}

@@

To illustrate projections with environmental effects we projected the best black sea bass model under three alternative scenarios where the AR1 time-series model for bottom temperature continues into the projection years, where we project with bottom temperature being the average of the most recent 5 years, and where bottom temperature increases following a predictions from a simple linear regression of the estimated bottom temperature anomalies over time.

@@

We found the best model of bottom temperature covariate effects included the effect only on the recruitment of the northern stock (Table \ref{tab:diff-aic-table}). However, the difference in AIC for the model that also included effects on recruitment for the southern component suggested some evidence for this hypothesis as well (Table \ref{tab:aic-wts-table}).

Across retrospective peels where the terminal year of data was sequentially reduced, the ranking for the best model was consistent. However, when the terminal year was reduced 6 or more years, there was more evidence for further effects of temperature on recruitment for the southern component and for temporal variation in natural mortality at age 1 (Table \ref{tab:aic-wts-table}).

The retrospective model fits did not indicate any evidence of patterns for northern or southern component SSB (all Mohn's $\rho \approx -0.03$) or fishing mortality (all Mohn's $\rho \approx 0.03$ and $-0.04$ for average $F$ at ages 6 and 7 in north and south regions, respectively) for any of the models (Table \ref{tab:rho-table}).

## Bottom temperature effects {-}

>>>>

A state-space statistical approach and maximum marginal likelihood or Bayesian fitting of stock assessment models allows estimation of time- and age-varying population attributes as random effects [@nielsenberg14,@cadigan16,@milleretal16]. This estimation approach is considered an essential feature of gold-standard assessment models that we use in tactical management of commercially important fish stocks [@puntetal20]. The State-space Assessment Model [SAM, @nielsenberg14] continues to be developed and remains widely used within ICES to assess European fish stocks. Various state-space models are being used to manage cod and plaice stocks in the waters of Eastern Canada [@varkeyetal22;@perreaultetal20] and to the south, the Woods Hole Assessment Model [WHAM, @millerstock20; @stockmiller21] is now used to assess many fish stocks in the Northwest Atlantic Ocean [@nefsc22;@nefsc22a;@nefsc24].

WHAM is an R package developed and maintained at NOAA's Northeast Fisheries Science Center [https://timjmiller.github.io/wham]. WHAM can be configured to fit a wide range of age-structured models from traditional statistical catch-at-age models without any random effects to models with several time and age varying process errors and effects of environmental covariates on various population parameters. Like SAM, WHAM models are built using the Template Model Builder package [TMB, @kristensenetal16] which provides a computationally efficient means of fitting an extremely wide class of models with random effects. WHAM has undergone active development since its creation and includes random effects options for recruitment, inter-annual transitions of numbers at age (hereafter referred to as "survival"), fishery and index selectivity, natural mortality, and catchability.

However, WHAM has only allowed one stock (component) and a single region and without seasonal changes in stock dynamics. Using such models for stocks that have subcomponents with varying seasonal movement can provide incorrect inferences and poor management advice [@bosleyetal22;@yingetal11;@caoetal14]. Furthermore, the ability to account for spatial structure and model multiple stocks are also important features of leading-edge assessment modeling frameworks [@puntetal20]. We describe here the implementation of these features and other extensions since @stockmiller21 in WHAM version 2.0. Many of these new configuration options can also be useful when modeling a single stock and region.

We developed this extension of WHAM in concert with a research track assessment process for black sea bass where new modeling frameworks were recently examined and modeling two stock components simultaneously and temporal changes in distribution was of interest [@nefsc23]. To illustrate the usage of WHAM 2.0 we apply it to two stock components of black sea bass off the coast of the NEUS using most of the model assumptions that were accepted during the peer-review process for the black sea bass assessment. However, here we evaluate evidence for alternative hypotheses of temporal variation and effects specifically of bottom temperature on recruitment and natural mortality of age 1 individuals.

When modeling multiple stock components occurring in different regions further dimensionality is added to the class of possible equilbrium reference points [e.g., @kapuretal21]. Short term projections are also an important part of the management process and the separation of observation and process errors in state-space models allows uncertainty in projected attributes such as recruitment and SSB without external simulation exercises.  Finally, the inclusion of explicit effects of environmental covariates in the model also requires some assumptions on how they are treated in projections and reference points. We demonstrate some of options of WHAM version 2.0 provide biological reference points and short-term projections.

@@

Many of the options and equations of WHAM version 2.0 are the same as those in @stockmiller21, so we will only describe extensions and differences that have occurred since their first description of WHAM. The new version of WHAM can model multiple stocks each with their own movement, harvest, and natural mortality. Much of the description below is for a specific stock $s$, but, for simplicity, this subscript is implicit except when necessary.

@@

As in previous versions of WHAM, the transitions in numbers at age from one year to another after recruitment can be treated deterministically or as functions of random effects. The predicted numbers at age in year $y$ at age $a$ for a given stock are vector analogs ($\mathbf{N}_{O,y,a}$) of the equations for numbers at age in the standard WHAM model [@stockmiller21]. For ages $a = 2,\ldots, A-1$, where $A$ is the plus group, the expected number alive in each region at the beginning of the following year and next age class age can be obtained from the first $n_R$ elements of the vector
$$\mathbf{P}_{y-1,a-1}' \mathbf{N}_{y-1,a-1}.$$
The numbers alive in each region can also be modeled more simply using the sub-matrix $\mathbf{O}_{y,a}$. The general model for the transitions in abundance at age is 
\begin{equation*}
\log\left(\mathbf{N}_{O,y,a}\right)|\mathbf{N}_{O,y-1,a-1} =  \log\left(\mathbf{O}_{y-1,a-1}' \mathbf{N}_{O,y-1,a-1}\right) + \boldsymbol{\varepsilon}_{y,a}
\end{equation*}
for ages $a = 2,\ldots, A-1$, and for the plus group
\begin{equation*}
\log\left(\mathbf{N}_{O,y,A}\right)|\mathbf{N}_{O,y-1,a-1},\mathbf{N}_{O,y-1,A} = \log\left(\mathbf{O}_{y-1,A-1}' \mathbf{N}_{O,y-1,A-1} + \mathbf{O}_{y-1,A}' \mathbf{N}_{O,y-1,A}\right) + \boldsymbol{\varepsilon}_{y,A}
\end{equation*}
where $\boldsymbol{\varepsilon}_{y,a}$ is the vector of region-specific errors for a given stock, which are independent across stocks and regions, but the same correlation structures as previous versions are possible across ages and years for a given stock and region.
When there is autocorrelation with age, WHAM now assumes this applies only to ages $a>1$ by default so that recruitment random effects are independent of those for for the annual transitions of older age classes. So the general covariance structure for a given stock at ages $a>1$ in region $r$ is that of a two-dimensional first-order autoregressive (2DAR1) process
\begin{equation*}
  Cov\left(\varepsilon_{y,a,r},\varepsilon_{y',a',r}\right) =   \frac{\rho_{N,\text{age},r}^{|a-a'|}\rho_{N,\text{year},r}^{|y-y'|}\sigma_{N,a,r}\sigma_{N,a',r}}{\left(1 -  \rho_{N,\text{age},r}^2\right)\left(1 - \rho_{N,\text{year},r}^2\right)} 
\end{equation*}
and that for age 1 is just AR1 across years 
\begin{equation*}
  Cov\left(\varepsilon_{y,1},\varepsilon_{y',1}\right) =   \frac{\rho_{N,1,\text{year}}^{|y-y'|}\sigma^2_{N,1}}{1 - \rho_{N,1,\text{year}}^2}.
\end{equation*}
When the transitions in abundance at age are treated deterministically, $\boldsymbol{\varepsilon}_{y,a} = 0$.  Since recruitment for a given stock currently only occurs in one region $r_s$ there is only a single time-varying recruitment random effect for each stock.

@@

For each season, there are at most $n_R-1$ parameters determining movement among regions given starting an the season in region $r$ in either the sequential or simultaneous configurations. Movement parameters are estimated on a transformed scale via a link function $g(\cdot)$. If survival and movement are occur simultaneously, the parameters are estimated a log link function is used and if they are separable, an additive logit link function (like a multinomial regression) is used. On the transformed scale, the general model for the movement parameter from region $r$ to $r'$ in season $i$ and year $y$ for individuals of age $a$ is a linear function of both random and environmental effects:
\begin{equation*}
  g(\mu_{r\rightarrow r',y,a,i}) = \theta_{r\rightarrow r',i} + \varepsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r \rightarrow r',a,i,k} E_{k,y}.
\end{equation*}
The random effects $\varepsilon_{r\rightarrow r',y,a,i}$ are season-, and(or) region-to-region-specific and modeled most generally as 2DAR1 random effects with age and(or) year where the covariance is
\begin{equation*}
  Cov\left(\varepsilon_{r\rightarrow r',y,a,i},\varepsilon_{r\rightarrow r',y',a',i}\right) =   \frac{\rho_{r\rightarrow r',\text{age},i}^{|a-a'|}\rho_{r\rightarrow r',\text{year},i}^{|y-y'|}\sigma^2_{r\rightarrow r',i}}{\left(1 -  \rho_{r\rightarrow r',\text{age},i}^2\right)\left(1 - \rho_{r\rightarrow r',\text{year},i}^2\right)}
\end{equation*}
similar to how WHAM models variation in survival, natural mortality, and selectivity. Effects of covariate $E_k$ can be age-, season-, and(or) region-to-region-specific $\beta_{r\rightarrow r',a,i,k}$ and the same orthogonal polynomial options in the previous versions of WHAM for effects on recruitment and natural mortality are available.

There is currently no likelihood component for tagging data. Therefore, movement parameters would generally either need to be fixed or assumed to have some prior distribution, possibly based on external parameter estimates. We include prior distributions for the season and region-to-region specific (mean) movement parameters which are treated as random effects with the mean defined by the initial value of the fixed effect counterpart and standard deviation
  \begin{equation*}
  \gamma_{r\rightarrow r',i} \sim \text{N}\left(\theta_{r\rightarrow r',i}, \sigma^2_{r\rightarrow r',i}\right).
  \end{equation*}
When priors are used, the movement is defined instead as
  \begin{equation*}
  g(\mu_{r\rightarrow r',y,a,i}) = \gamma_{r\rightarrow r',i} + \varepsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r\rightarrow r',a,i,k} E_{k,y} 
  \end{equation*}

@@

Natural mortality options have been expanded in WHAM. When not estimated, (mean) mortality rates may be stock-, region-, and age-specific. When random effects are used, the same 2DAR1 structure with age and year as described by @stockmiller21 can be configured for a given stock and region. Any environmental covariate effects can be stock-, region-, and age-specific. So the general model for natural mortality is
\begin{equation*}
  \log M_{y,a,r} = \theta_{M,r} + \varepsilon_{M,r,y,a} + \sum^{n_E}_{k=1} \beta_{M,r,a,k} E_{k,y}.
\end{equation*}
The general covariance structure for random effects are modeled most generally as 2DAR1 random effect with age and(or) year where the covariance is
\begin{equation*}
  Cov\left(\varepsilon_{M,y,a,r},\varepsilon_{M,y',a',r}\right) =   \frac{\rho_{M,\text{age},r}^{|a-a'|}\rho_{M,\text{year},r}^{|y-y'|}\sigma^2_{M,r}}{\left(1 -  \rho_{M,\text{age},r}^2\right)\left(1 - \rho_{M,\text{year},r}^2\right)}.
\end{equation*}

@@

Catchability of the index can also be treated as functions of normal random effects and(or) environmental covariate effects 
$$log \frac{q_{m,y}-l_m}{u_m-q_{m,y}} = \theta_{q,m} + \varepsilon_{q,m,y}  + \sum^{n_E}_{k=1} \beta_{q,m,k} E_{k,y}$$ 
where $u_{m}$ and $l_{m}$ are the upper and lower bounds of catchability for index $m$ (defaults are 0 and 1000) and the general covariance structure for the annual random effects is AR1
$$Cov\left(\varepsilon_{q,m,y},\varepsilon_{q,m,y'}\right) =   \frac{\rho_{q,m}^{|y-y'|}\sigma^2_{q,m}}{1 - \rho_{q,m}^2}.$$

@@

For $X\%$ SPR-based reference points, we use a Newton method and iterate
\begin{equation}\label{eq:newton}
  \log\widetilde{F}^{(i)} = \log\widetilde{F}^{(i-1)} - \frac{g\left(\log\widetilde{F}^{(i-1)}\right)}{g'\left(\log\widetilde{F}^{(i-1)}\right)}
\end{equation}
where $g(\log F)$ is the difference between the weighted sums of spawning biomass per recruit at $F$ and $X$\% of unfished spawning biomass per recruit across stocks:
\begin{equation}\label{eq:newton-obj}
  g(\log F) = \sum^{n_s}_{s=1} \lambda_s\left[\upphi_s\left(F = e^{\log F}\right) - \frac{X}{100}\upphi_s\left(F=0\right)\right].
\end{equation}
where $\upphi_s\left(F=0\right)$ is the equilibrium unfished spawning biomass per recruit. $g'(\log F)$ is the derivative of $g$ with respect to $\log F$, and the weights to use for each stock $\lambda_s$ can be specified by the user or relative to the average of recruitment for each stock over the same years the user defines to calculate ``static'' equilibrium spawning biomass and yield.

@@

The projection options are generally the same as those for previous versions of WHAM. When there is movement of any stocks, the user has the option to project and use any random effects for time-varying movement or use the average over user specified years, analogous to how natural mortality can be treated in the projection period. The projection of any environmental covariates has been revised to better include error in the estimated latent covariate in any effects on the population in projection years. Users can also specify a catch or fully-selected fishing mortality in each projection year and they can be fleet-specific.

@@

Black sea bass are a temperate reef fish in the western Atlantic Ocean ranging from the Gulf of Mexico to the Gulf of Maine.  Fish north of Cape Hatteras, NC, are considered to comprise a single stock unit, but individual populations exhibit spawning site fidelity
[@ablehales97;@fabrizioetal13]. Fish in this northern stock unit perform seasonal migrations out on the continental shelf in the fall and back to their inshore spawning areas during the spring [@musickmercer77]. Analyses of tagging studies have demonstrated that the extent of seasonal migration varies along the coast such that fish from populations off of southern New England and further north move offshore and as far south as the coasts of Virgina and North Carolina, whereas fish in the southern portion of the stock range generally move shorter distances between inshore and offshore areas of similar latitude [@mosershepherd09].

@@

Leading up to the 2023 peer-reviewed assessment, a working group (hereafter referred to as "Working Group") composed of scientists from federal, state, and academic institutions determined an optimal data and model configuration for the black sea bass stock using the multi-stock and multi-region extension of WHAM described above [@nefsc23]. This assessment included the spatial features and investigated inclusion of hypothesized environmental drivers that were prioritized research recommendations from previous black sea bass assessments.

Below we describe the assumptions and configuration of the assessment model as determined by the Working Group as well as the alternative assumptions for recruitment and natural mortality in the models we fit to evaluate alternative hypotheses of bottom temperature effects on black sea bass.

@@

The two monthly movement matrices are
\begin{equation*}
\boldsymbol{\mu}_{1} = 
  \begin{bmatrix}
     1-\mu_{\text{N}\rightarrow \text{S}} & \mu_{\text{N}\rightarrow \text{S}} \\
     0 & 1 \\
  \end{bmatrix}
\end{equation*}
for the portion of the year after spawning and
\begin{equation*}
\boldsymbol{\mu}_{2} = 
  \begin{bmatrix}
     1 &  0 \\
     \mu_{\text{S}\rightarrow \text{N}} & 1-\mu_{\text{S}\rightarrow \text{N}} \\
  \end{bmatrix}
\end{equation*}
for the portion of the year before spawning. As noted in the description of the general WHAM model, tagging data are not yet allowed. However, the Working Group also fit a Stock Synthesis model [@methotwetzel13] which provided estimated movement rate parameters and standard errors that were used to configure the priors for WHAM (see Supplementary Materials).

@@

## Uncertainty in recreational CPA index observations {-}

The estimated coefficients of variation (CVs) provided by the analyses to generate the recreational catch per angler (recreational CPA) index ranged between 0.02 and 0.06 which the Working Group felt did not capture the true observation uncertainty in the index with regard to its relationship to stock abundance. In many of the initial runs as well as the base model we allowed a scalar multiple of the standard deviation of the log aggregate index to be estimated for these indices in the northern and southern regions. Models that successfully estimated these scalars indicated standard deviations for these surveys to be  approximately 5 times the input value and this value was fixed in many preliminary runs to avoid dealing with convergence problems. However, the base model successfully estimated these scalars. The model estimates are negligibly affected by estimating these scalars, but we felt estimating these parameters allowed uncertainty in model output to be more properly conveyed.

@@

The Working Group investigated many alternative assumptions for the probability models and selectivity models for the 8 different sets of age composition observations to reduce residual patterns and retrospective patterns. These analyses resulted in use of selectivity random effects for the northern fleet and indices and logistic-normal likelihoods for 6 sets of of age composition observations and Dirichlet-multinomial likelihoods for one index and one fleet in the northern region (Table \ref{tab:age-comp-sel-table}).

@@

\begin{equation}\label{eq:expected-recruitment}
E\left(\log N_{y,1}|x_y\right) = \mu_{R} + \beta_{R} x_y.
\end{equation}
Similarly, expected log-natural morality as a function of bottom temperature is
\begin{equation}\label{eq:expected-M1}
E\left(\log M_{y,1}|x_y\right) = \mu_{M,1} + \beta_{M} x_y
\end{equation}
Because age 1 fish for the northern component can exist in both regions after January 1, natural mortality is acting in both regions for this stock. For models with covariate and/or annual random effects for age 1 fish we assume them only in the northern regions for the northern component. The corresponding random effects are
\begin{equation}\label{eq:Rec-re}
\log N_{y,1} = E\left(\log N_{y,1}|x_y\right) + \varepsilon_{y,1}
\end{equation}
and
\begin{equation}\label{eq:M-re}
\log M_{y,1} = E\left(\log M_{y,1}|x_y\right) + \varepsilon_{M,y}.
\end{equation}
The constant or mean log-natural mortality rate is assumed to be $\mu_{M,a} = \log(0.4)$ for all ages as recommended by the Working Group. Because the bottom temperature anomalies and random effects are centered at 0, mean log-natural mortality at age 1 over the time series should be approximately equal to $\mu_{M,1}$ for models that include those effects.

## Model fitting, diagnostics, BRPs, and projections {-}

@@

We produced SPR-based BRP estimates ($F_{40\%}$ and SSB$_{40\%}$) from the best performing model using a variety of options provided in WHAM. We generated annual BRP estimates based on the annual weight at age, maturity at age, M, fleet selectivity, and either the annual recruitment random effect (Eq. \ref{eq:Rec-re}) or the annual expected recruitment (Eq. \ref{eq:expected-recruitment}). We also calculated status as the ratios of annual average F and SSB to the corresponding reference points. Finally, we calculated "prevailing" BRPs using the average weight at age, maturity at age, M, fleet selectivity from the last 5 years (2017 to 2021) and average recruitment over years 2000 to 2021, and made Kobe plots of joint status (Kobe plots) for the terminal year of the model (2021). All BRPs and status estimates are generated for each region as well as for the total stock.

We projected the best black sea bass model under three alternative scenarios for the bottom temperature covariate where 1) the AR1 time-series model continues into the projection years, 2) the average of the most recent 5 years is projected, and 3) bottom temperature increases in projection years following a prediction from a simple linear regression of the estimated bottom temperature anomalies over time. We specified fishing mortality in projection years to be constant at the value in the last model year (2021). Average weight, natural mortality, movement, and maturity at age are used in projection years whereas the AR1 processes for all numbers at age random effects are continued.

@@

We found the best model of bottom temperature covariate effects included the effect only on recruitment of the northern stock and no random effects on age 1 natural mortality ($M_1$, Tables \ref{tab:diff-aic-table} and  \ref{tab:aic-wts-table}). Model $M_1$ consistently had the lowest AIC across retrospective peels where the terminal years of data are removed from the model. The difference in AIC for the model that also included bottom temperature effects on recruitment for the southern component ($M_3$) suggested some evidence for this hypothesis as well. There was little evidence for also including random effects on age 1 natural mortality for the northern stock component. However, when more than three terminal years of data were removed, the evidence for adding age 1 natural mortality random effects for the northern component ($M_8$) increased is similar to adding bottom temperature effects on southern component recruitment ($M_3$).
